---
title: "Olink® PEA Functional Enrichment Analysis"
author: "NICHD Bioinformatics and Scientific Programming Core"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 3
params:
  npx_data:        NULL
  sampletable:     NULL
  data:            NULL
---

# Analysis Overview

This report performs functional enrichment analyses (ORA and/or GSEA, depending
on your config) on the **primary effects** from your Olink PEA differential
abundance results.

With shipped **test data**, there is only one primary contrast:

- `disease_vs_control` (Group: disease vs control, adjusted for Cov1 and Cov2)

If you are running real data with additional contrasts, update the
**USER SETTINGS** section to choose which contrasts to enrich.

```{r AnaylzeOlink, eval=FALSE, results='hide'}
# When running interactively, and not from the Snakefile,
# load the AnalyzeOlink R package, which is stored locally.
# This package has many custom functions used throughout this document.
devtools::document('../../AnalyzeOlink')
devtools::load_all('../../AnalyzeOlink')
```

```{r setup, include=FALSE}
library(dplyr)
library(purrr)
library(AnalyzeOlink)

# Read config
config <- yaml::read_yaml("../config/config.yaml")

# Set up cache invalidation
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  cache.extra = list(
    file.info("../config/config.yaml")$mtime,
    file.info(config$input_data_paths$PEA)$mtime,
    file.info(config$sampletable_paths$PEA)$mtime,
    file.info(config$output_data_paths$PEA)$mtime
  )
)

# Use Snakemake inputs if running via Snakefile
# otherwise, fall back to config contents
if (!is.null(params$npx_data)) {
  config$input_data_paths$PEA <- params$npx_data
}
if (!is.null(params$sampletable)) {
  config$sampletable_paths$PEA <- params$sampletable
}
if (!is.null(params$data)) {
  config$output_data_paths$PEA <- params$data
}
```

```{r fetch-data-from-olinkanalyze, results='asis', eval=TRUE, cache=TRUE}
olink_analyze_data <- readRDS(config$output_data_paths$PEA)
se_list            <- olink_analyze_data$se_list
res_list           <- olink_analyze_data$res_list
```

```{r subset-res-list-primary-effects, eval=TRUE, cache=TRUE, dependson='fetch-data-from-olinkanalyze'}
# Keep statistics for primary effects only (Group for disease_vs_control)
res_list_subset_1_effect <- keep_primary_effects(res_list)
```

# Enrichment Settings (USER-CONFIGURABLE)

Most users only need to edit this block.

```{r enrichment-user-settings, cache=FALSE}
# -------------------------------------------------------------------------
# USER SETTINGS
# -------------------------------------------------------------------------

# 1) Which contrasts should be enriched?
#    - With test data: only disease_vs_control exists.
#    - With real data: pick any subset you care about.
contrasts_to_enrich <- names(res_list_subset_1_effect)

# 2) OPTIONAL: override enrichment methods here if you want to temporarily
#    change what is in config$enrichment without editing the YAML.
#    Example: method_override <- c("ORA", "GSEA")
method_override <- NULL

# -------------------------------------------------------------------------
# Sanity checks
# -------------------------------------------------------------------------
bad_ctr <- setdiff(contrasts_to_enrich, names(res_list_subset_1_effect))
if (length(bad_ctr) > 0) {
  stop("Unknown contrasts in contrasts_to_enrich: ", paste(bad_ctr, collapse = ", "))
}

# Subset to chosen contrasts
res_list_subset_1_effect <- res_list_subset_1_effect[contrasts_to_enrich]

# Temporarily override config methods if requested
if (!is.null(method_override)) {
  config$enrichment$methods <- method_override
}
```

```{r functional-enrichment, eval=TRUE, cache=TRUE, dependson=c('subset-res-list-primary-effects','enrichment-user-settings')}
# Perform functional enrichment analysis on all chosen contrasts
enrich_list <- if (isTRUE(config$parallel)) {
  parallel_enrich(res_list_subset_1_effect, config)
} else {
  enrich(res_list_subset_1_effect, config)
}
```

## Functional Enrichment Results {.tabset}

Enrichment tables list gene set names ranked by representation metrics. The
following abbreviated prefixes in the Description column denote the source
database or publication:

- **GOMF**: "Gene Ontology - Molecular Function"
- **GOBP**: "Gene Ontology - Biological Process"
- **GOCC**: "Gene Ontology - Cellular Component"
- **Reactome**: curated pathway database
- **KEGG**: Kyoto Encyclopedia of Genes and Genomes
- **LIU**, **CHEN**, **RUTELLA**, etc.: custom gene sets from published studies

```{r functional-enrichment-results, results='asis', eval=TRUE, cache=FALSE, dependson='functional-enrichment'}
# NOTE: rendering from cache will cause DT tables to disappear
print_enrichment_results(enrich_list, config)
```

## Dotplots {.tabset}

ORA dotplots show top enriched gene sets by RichFactor (x) and adjusted p-value
(color). Point size indicates overlapping gene count.

GSEA dotplots show top gene sets by NES (x). Color and size are as above.

```{r dotplots, results='asis', eval=TRUE, cache=FALSE, dependson='functional-enrichment'}
dotplot(enrich_list, config)
```

## Session Information

```{r sessioninfo, eval=TRUE}
sessionInfo()
```
